---
title: "Shock-Recovery (DV3) Variables — Descriptive Statistics"
subtitle: "NYS cities · GFC (2009) and COVID (2020) · drawdown depth + time-to-recover"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    fig-align: center
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr); library(readr); library(gt); library(here); library(knitr)

tab <- function(name) here::here("outputs", "recovery_descriptives", "tables", paste0(name, ".csv"))
fig <- function(name) here::here("outputs", "recovery_descriptives", "figures", paste0(name, ".png"))

# Render a saved CSV as a gt table, formatting numeric columns.
show_tbl <- function(name, title = NULL, subtitle = NULL, decimals = 3) {
  d <- readr::read_csv(tab(name), show_col_types = FALSE)
  num <- names(d)[vapply(d, is.numeric, logical(1))]
  pct <- intersect(c("pct_missing", "na_baseline", "recovered_pct", "censored_pct",
                     "notrecovered_pct"), names(d))
  g <- gt::gt(d)
  if (!is.null(title)) g <- gt::tab_header(g, title = title, subtitle = subtitle)
  if (length(num)) g <- gt::fmt_number(g, columns = dplyr::all_of(setdiff(num, c(pct, "n", "shock"))), decimals = decimals)
  if (length(pct)) g <- gt::fmt_percent(g, columns = dplyr::all_of(pct), decimals = 1)
  if ("n" %in% names(d)) g <- gt::fmt_number(g, columns = "n", decimals = 0)
  if ("shock" %in% names(d)) g <- gt::fmt_number(g, columns = "shock", decimals = 0, use_seps = FALSE)
  g
}
```

This report describes the **DV3 shock-recovery** variables — the *bounce-back* facet of fiscal
resilience that the stability measures (DV2) do not capture. The grain is an **event study**:
one row per **entity × shock × target series**, for the two common shocks **2009 (GFC)** and
**2020 (COVID)**. Sample: NYS **cities** (NYC absent from the OSC data). Three target series are
tracked: **reserves** (`available_fb_ratio`, available fund balance ÷ GF expenditure — a *level*
drawdown in ratio points, undeflated), **own-source revenue** and **operating expenditure**
(real-dollar *proportional* drawdowns, CPI-U deflated to 2023). Artifacts are produced by
`code/65_recovery_descriptives.R`; this document presents them. *Lower drawdown + faster recovery
= more resilient.*

## 1. Definitions

For each unit and shock year `t0`, a pre-shock **baseline** `B` is the mean of the target over
`[t0-3, t0-1]`. **Drawdown** is the worst shortfall below `B` over `[t0, t0+6]`; **recovery time**
is the first year the target climbs back to `B`. Because raw fund balance is near-zero/negative
for some cities, reserves use a **level (ratio-point)** drawdown rather than a proportional one
(which would explode); the dollar flows use the proportional form.

```{r}
show_tbl("01_definitions")
```

## 2. Construction

Targets are CPI-U deflated **before** the metrics where they are dollar flows, so neither
inflation nor the ARPA-driven nominal fund-balance surge is mistaken for recovery; the reserves
ratio is already real and is not deflated. Shocks `t0 = 2009, 2020`; window 6 years; horizon 4
years. CPI-U provenance: `transparency/reference_sources.md`.

```{r}
show_tbl("02_construction")
```

## 3. Descriptive statistics & distribution

Drawdown scale differs by series (reserves in **ratio points**, flows **proportional**), so read
each series on its own row. `recovery_years` is summarized over **recovered** cases only.

```{r}
show_tbl("03_summary_stats", decimals = 3)
```

Recovery status shows the headline asymmetry between the two shocks: the **GFC window is fully
observed** (no censoring), while **COVID recovery is heavily right-censored** — most clearly for
operating expenditure (≈ 52% censored), because the panel ends before the post-COVID window
closes. So COVID is a *depth* story; its recovery *speed* is largely unknown.

```{r}
show_tbl("03_recovery_status")
```

```{r}
#| fig-cap: "Drawdown-depth distributions by series and shock (winsorized 1/99 within series)."
knitr::include_graphics(fig("03_drawdown_distributions"))
```

## 4. By shock (GFC vs COVID)

There is no annual time series at this grain; the comparison that matters is **2009 vs 2020**.
Two things stand out: **operating spending barely dips** in either shock (the sticky-spending
finding again — services are protected), and **reserves were hit harder by the GFC than by
COVID** (mean drawdown ≈ 0.13 vs 0.08 ratio points) — consistent with federal relief (ARPA)
cushioning the COVID reserve hit. Labels show the share that recovered within the window.

```{r}
#| fig-cap: "Mean drawdown depth by series and shock; labels = share recovered within the window."
knitr::include_graphics(fig("04_by_shock"))
```

## 5. Depth vs speed

Do deeper shocks take longer to climb back? Among **recovered** cases, drawdown depth and
recovery time are **positively correlated for every series** (Spearman ≈ 0.40–0.63) — deeper
reserve/revenue drawdowns rebuild more slowly, strongest for own-source revenue. This is the
substantive content of DV3 that a stability measure misses.

```{r}
show_tbl("05_depth_vs_speed", decimals = 2)
```

```{r}
#| fig-cap: "Years to recover vs drawdown depth (recovered cases only), by series and shock."
knitr::include_graphics(fig("05_depth_vs_speed"))
```

## 6. Distribution by size quintile

Size class = within-city quintile of time-averaged operating expenditure (1 = smallest).
Drawdown depth winsorized within series.

```{r}
show_tbl("06_by_size_quintile", decimals = 3)
```

```{r}
#| fig-cap: "Drawdown depth by size quintile, by series (winsorized within series)."
knitr::include_graphics(fig("06_by_size_quintile"))
```

---

*Reproduce:* run `code/40_construct_resilience.R` (builds `<entity>_recovery.rds`) then
`code/65_recovery_descriptives.R` to (re)generate the artifacts, then
`quarto render analysis/recovery_descriptives.qmd`.
